---
title: "R_programming"
author: "REACH"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = F, message = F)
```

# {.tabset}

## Session 1 

### why R

R is a scripting language known for its simple syntax.
It is suitable for anyone desiring start or enhance their journey in
data management, analysis and scientific research. It can be used to
retrieve data, clean, analyse and visualize the data.

- Simple syntax
- Variety of packages to handle wide range of tasks
- Powerful RStudio IDE for simplify code management and getting started
- Also R has a large community of users and you will easily get support

### Installing R and RStudio

Before we start coding, we need to setup the coding environment. 
We shall install R and RStudio the graphical Integrated Development Environment for R.
RStudio makes using R much easier and more interactive.

### Installing R

Download the R installation file from **[Download](https://cran.r-project.org/bin/windows/base/)**.

Run the downloaded file and accept prompts to install.

Download the R installation file from **[Download](https://cran.r-project.org/bin/windows/base/)**.

Run the downloaded file and accept prompts as well.

### RStudio Interface



```{r }

```

## Session 2


```{r }

```

## Session 3


```{r }

```
