---
title: "week_2"
author: "REACH"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float:
      collapsed: false
      smooth_scroll: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = F, message = F, results = "hide")
```

# Reading and manipulating data

In order to manipulate data, we need to create, read/import data into the R environment using appropriate functions.
We can use functions provided in the base system where applicable or functions in other containers (packages) specialized for specific file types. 

## Extending R base functionality with packages

### What are packages
R comes with functionality in the base system. This functionality is what you get when you download and install R.
This can be extended using containers (packages) that have other functionality or different workflow. These packages contain functions to perform specific tasks.

### Package installation

For packages on The Comprehensive R Archive Network (CRAN), you can install them using the **install.packages()** function, specifying the name of the function in quotes. This can be done in the console pane since we do not need to save this in a script.

> The code below will install all tidyverse related packages.

```{r}
install.packages("tidyverse")
```

**NOTE** For packages under development by developers that may not be on CRAN, we can install them using the **devtools** or **remotes** packages. Let us install the **devtools** package. With **install.packages("devtools")**.
We can now install a package using devtools

```{r}
install.packages("devtools")

devtools::install_github("twesigye10/supporteR")

```

### Package loading

We can load packages using the **library()** function. 

```{r}
library(tidyverse)
library(supporteR)

```







## Reading and writing files


## Manipulating data


## Pipes

