---
title: "week_2"
author: "REACH"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float:
      collapsed: false
      smooth_scroll: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(eval = FALSE, echo = TRUE, warning = F, message = F, results = "hide")
```

# Getting started with R Markdown

Rmarkdown helps us to generate reports. The reports could be based on changing data like during data collection where you have to communicate progress to different teams. It could also be a dynamic report generated regularly after a given period of time like a weekly or monthly report. when we connect these reports with data, it becomes easy for us to run the code and generate these reports.

These reports can include different elements like text, tables, graphs, maps e.t.c that are based on the data connected to the reports. 

The reports can be generated in different formats like HTML files, PDF files, Word files, Presentation files ...

## Creating and an Rmarkdown file



## Basic parts of the file

## Basic syntax

headings
normal text
- links
code chunks
inline code

## Generate the report


- https://rmarkdown.rstudio.com/authoring_basics.html
- https://monashbioinformaticsplatform.github.io/2017-11-16-open-science-training/topics/rmarkdown.html




```{r}

```

code shortcuts