---
title: "week_2"
author: "REACH"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float:
      collapsed: false
      smooth_scroll: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(eval = FALSE, echo = TRUE, warning = F, message = F, results = "hide")
```

# Creating Graphs using ggplot2

We can create graphs in R using different packages. For the purpose of our learning, we shall use **[ggplot2](https://ggplot2.tidyverse.org/reference/)** from tidyverse. This package helps us to quickly create beautiful graphs using data in the data frame and we can easily customize these graphs according to our preference.

From the reference, you can click on the icons of desired graphs under the `Geoms` sub heading to find more about the graph and how to create it.

There is also another resource of **[R Graph Gallery](https://r-graph-gallery.com/ggplot2-package.html)** on ggplot2 where you can access different graphs and explore how to create and customize them.

ggplot graphs are built step by step, incrementing each step with a `+` sign at the end of a step.

## Basic steps for creating graphs

- Attach data frame to the ggplot using the data argument
- Specify the mappings/aesthetics (aes). Columns and other properties to visualize
- Specify the type of plot/graph by adding the geo_*() functions.

## Bar graph

We use the `geom_bar()` or `geom_col()` functions to create the bar graphs. `geom_bar()` uses `stat_count()` by default and makes the height of the bar proportional to the number of cases in each group. `geom_col()` uses `stat_identity()` and the heights of the bars represent values in the data.

```{r}
library(tidyverse)

df_iris <- read_csv("inputs/iris_data.csv") |> 
  mutate(sepal_length_category = case_when(Sepal.Length < 5 ~ "cat_less_than_5",
                                                          Sepal.Length < 7 ~ "cat_5_6",
                                                          Sepal.Length >= 7 ~ "cat_7+"))

# basic geom_bar
ggplot(data = df_iris, aes(y = Species)) + # x or y column provided
  geom_bar(fill = "blue") + # uses stat_count() by default
  theme_bw()

# geom_bar color by category
ggplot(data = df_iris, aes(x = Species)) + # x or y column provided
  geom_bar(aes(fill = sepal_length_category)) + # uses stat_count() by default
  theme_bw()

# basic geom_col
ggplot(data = df_iris, aes(x = Sepal.Length, y = Species)) + # both x and y provided
  geom_col(fill = "blue") +  # uses stat_identity() by default
  theme_bw()

# geom_col color by category
ggplot(data = df_iris, aes(x = Sepal.Length, y = Species)) + # both x and y provided
  geom_col(aes(fill = sepal_length_category)) +  # uses stat_identity() by default
  theme_bw()

# summarizing data and plotting
df_mean_sep_len <- df_iris |> 
  select(Sepal.Length, Species) |> 
  group_by(Species) |> 
  summarise(`Mean Length` = mean(Sepal.Length))

# customizing the graph
ggplot(data = df_mean_sep_len, aes(x = `Mean Length`, y = Species)) +
  geom_col(fill = "blue") +
  theme_bw() +
  theme(axis.ticks = element_blank(),
        axis.text.x = element_text(face = "bold", size=12),
        axis.text.y = element_text(face = "bold", size=12),
        panel.grid.major.x = element_blank(),
        panel.grid.minor.x = element_blank(),
        panel.grid.minor.y = element_blank(),
        panel.border = element_blank(),
        axis.title = element_text(size=12)) +
  ylab(label = "")  

```

## Scatter plot

The scatter plot helps us to analyse the relationship between observations. It can be handy during Exploratory Data Analysis (EDA) to explore the data.

```{r}

library(tidyverse)

df_iris <- read_csv("inputs/iris_data.csv")

# Basic scatter
ggplot(data = df_iris, aes(x = Sepal.Length, y = Sepal.Width)) +
    geom_point() +
    labs(title = "Plot of Sepal measurements")

# scatter with color categories
ggplot(data = df_iris, aes(x = Sepal.Length, y = Sepal.Width, colour = Species)) +
    geom_point() +
    labs(title = "Plot of Sepal measurements")
```


## Boxplot

The Boxplot can help us to check the distribution of the data

```{r}
library(tidyverse)

df_iris <- read_csv("inputs/iris_data.csv")

# basic Boxplot
ggplot(data = df_iris, aes(x = Species, y = Sepal.Length)) +
    geom_boxplot()

# Boxplot with jitter
ggplot(data = df_iris, aes(x = Species, y = Sepal.Width, colour = Species)) +
    geom_boxplot() +
  geom_jitter()
```

## Export plots

We can use ggsave to export graphs to use in presentations and reports. It uses the defaults of the last plot displayed to export your plot.

```{r}
ggsave("outputs/plot.png", scale = 2)
```

